---
title: 'Assignment #1 (formative)'
author: "Michelle Papadakis Barradas"
date: "This assignment is due Thursday 12 October 2023 at 4pm"
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = FALSE) 
```

## Exercise 1
### Forking a repository 

"My forked version of the assignment template repository can be found at: https://github.com/MichellePapadakis/472_assignment_template" 


## Exercise 2
### Tidy data
The data from the built-in dataset used for this exercise was *HairEyeColor* which **is** _tidy_. We can verify this by examining the structure of the built-in dataset *HairEyeColor* and confirm that it complies to the three principles of a _*tidy data*_ which are:

1. Each value occupies its own cell. 
2. Each observation corresponds to its own row. 
3. Each variable is allocated its own column.

We could transform our data frame and maintain the _tidy version_  by pivoting it wider adding two columns, each for each sex. This transformation enables a better visualization by gender.


To _*untidy*_ the data we could pivot the original data frame into a longer version. In this case we could pass the information contained in the coumns *Hair* and *Eyes* to one column named *Attribute* and assign the values to a column named *Colour*. In summary I merge two variables, which originally had their own separate columns, into a single column called 'attributes. The values were grouped under the 'colours' column. This transformation resulted in an untidy format for the data frame.

```{r exercise # 2, warning=FALSE, message=FALSE}
#To work with the built-in data frames and table transformations, the tidyr package will be used. If the package is not installed, the following line should be uncommented.
#installed.packages("tidyverse")
library("tidyverse")
#Looking at built-in dataset from R, in this the data set is called "HairEyeColor"
data("HairEyeColor")
#Transforming the data set into Tibbles
HEC <- as.tibble(HairEyeColor)
#Pivoting wider to better fit the data 
HECw <- pivot_wider (HEC, 
                   names_from = "Sex", 
                   values_from = "n")
#Pivot longer to untidy de data, in this case we merge the two variables Hair and Color that previously had their own separate columns into one column named Attributes and created another called Colour to store the values.

HECl <- HEC %>%
  pivot_longer(cols = c(Hair, Eye),
               names_to = "Attribute",
               values_to = "Colour")
```
``````{r, echo=TRUE}
#Figure 1, original tidy
head(HEC)
#Figure 2, wide tidy
head(HECw)
#Figure 3, long untidy
head(HECl)
```

## Exercise 3
### Transforming data


```{r, exercise 3, echo=TRUE}

#Calling a built in data set and transforming it to tibble to better manipulation
data("population")
population_tibble <- as.tibble(population)
head(population_tibble)

#Converting the 'population' data frame into a tidy format, where each year has its own column, eliminating the need to repeat the country name for each year's data.
population_tibble_w <- population_tibble %>%
  pivot_wider(names_from = year, values_from = population)
head(population_tibble_w)

#To calculate the mean population for each country over the span of 19 years, we can approach it as a two-step process. First, we compute the total population for each year across all countries. Subsequently, we employ a mutation operation to divide the cumulative population total by the number of observations to get the mean population (in this case because we wanted to transform a numeric variable after having called summarise() according the instructions.  
 
exercise_3 <- population_tibble_w %>%
  group_by(country) %>%
  summarise(sumPop_1995to2013 = sum(c_across(c("1995", "1996", "1997", "1998", "1999", "2000", "2001", "2002", "2003", "2004", "2005", "2006", "2007", "2008", "2009", "2010", "2011", "2012", "2013")), na.rm = TRUE)) %>%
 mutate(meanPop_1995to2013 = ((sumPop_1995to2013)/19))
head(exercise_3)

```

## Appendix: All code in this assignment

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE} 
{r knitr::opts_chunk$set(warning = FALSE, message = FALSE)}
```
